---
title: 'Longitudinal Data: Repeated Measures'
subtitle: 'Mixed Models 1'
author: Alan Hubbard
date: Oct 25, 2018
output: 
  beamer_presentation:
    slide_level: 2  
header-includes:
  - \usepackage{pdfpages}
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
# This bit of code lets you set the font size
# in heading of code junk when writing to beamer.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
# Libraries
library(dplyr)
library(tidyverse)
library(tidyr)
library(lme4)
```

## Read in data
```{r prelim1, size="tiny"}
dat1=read_csv("deeks_ex1.csv")
tbl_df(dat1)
```

# Binary time-independent variable (balanced data)
## Random Intercept Model
```{r ex1a, size="tiny"}
# random intercept
lme_1 <- lmer(cd4 ~ binage + (1|id), data=dat1)
summary(lme_1)
```

## post-estimation stats
```{r ex1stats, size="tiny"} 
library(sjstats)
# ICC - var(b0i)/(var(b0i)+var(eij))
icc(lme_1)
AIC(lme_1)
# Now do by grabbing relevant objects
rand_vars1 <- re_var(lme_1)
rand_vars1
rand_vars1[2]/(rand_vars1[1]+rand_vars1[2])
```

## Random intercept and slope model

```{r ex1b, size="tiny"}
# random slope - try it, it won't work (not identifiable, why)
#lme_1b <- lmer(cd4 ~ (binage|id), data=dat1)
```

# Binary time-dependent variable (balanced data)
## Read in data
```{r prelim2, size="tiny"}
dat2=read_csv("deeks_ex2.csv")
tbl_df(dat2)
```

## Random Intercept Model
```{r ex2a, size="tiny"}
# random intercept
lme_2 <- lmer(cd4 ~ medvl + (1|id), data=dat2)
summary(lme_2)
```

## post-estimation stats
```{r ex2stats, size="tiny"} 
# ICC - var(b0i)/(var(b0i)+var(eij))
icc(lme_2)
AIC(lme_2)
# Now do by grabbing relevant objects
rand_vars2 <- re_var(lme_2)
rand_vars2
rand_vars2[2]/(rand_vars2[1]+rand_vars2[2])
```

## Random intercept and slope model
```{r ex2b, size="tiny"}
# random slope also not identifiable here
#lme_2b <- lmer(cd4 ~ (medvl|id), data=dat2)
```

# Continuous time-dependent variables (very unbalanced data)

## Read in data
```{r prelim3, size="tiny"}
dat3=read_csv("deeks_ex3.csv")
tbl_df(dat3)
```

## Random Intercept Model
```{r ex3a, size="tiny"}
# random intercept
lme_3 <- lmer(cd4 ~ logvlbase+logvlchange + (1|id), data=dat3)
summary(lme_3)
```

## post-estimation stats
```{r ex3astats, size="tiny"} 
# ICC - var(b0i)/(var(b0i)+var(eij))
icc(lme_3)
AIC(lme_3)
# Now do by grabbing relevant objects
rand_vars3 <- re_var(lme_3)
rand_vars3
```

## Random intercept and slope model
```{r ex3b, size="tiny"}
# random slope also not identifiable here
lme_3b <- lmer(cd4 ~ logvlbase+logvlchange+(logvlchange|id), data=dat3)
summary(lme_3b)
```


## post-estimation stats
```{r ex3bstats, size="tiny"} 
icc(lme_3b)
AIC(lme_3b)
rand_vars3b <- re_var(lme_3b)
rand_vars3b
```

## Reporting the estimated distribution of associations from random coefficients model.
```{r rand_coef, size="tiny", echo=F}
b2hat <- fixef(lme_3b)[3]
sdb2i <- sqrt(rand_vars3b[3])
b2i <- seq(b2hat-3*sdb2i,b2hat+3*sdb2i,length=100)
dist.b2i <- tibble(b2i,dens=dnorm(b2i,b2hat,sdb2i))
ggplot(dist.b2i,aes(x=b2i,y=dens))+geom_line()+geom_vline(xintercept = b2hat)+geom_vline(xintercept = 0)+scale_x_continuous(breaks = seq(-150,50,25))

```

## Which model does one choose (3 or 3b)?
```{r aic_comp, size="tiny"}
AIC(lme_3,lme_3b)
```
- The random intercept and slope model has lower AIC and thus fits the data relatively better than the random intercept model alone.


