---
title: 'Longitudinal Data: Repeated Measures'
subtitle: 'Mixed Models 2'
author: Alan Hubbard
date: Oct 25, 2018
output: 
  beamer_presentation:
    slide_level: 2  
header-includes:
  - \usepackage{pdfpages}
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
# This bit of code lets you set the font size
# in heading of code junk when writing to beamer.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40),tidy=TRUE)
# Libraries
library(tidyverse)
library(lme4)
library(sjstats)
```

## Read in data
```{r prelim1, size="tiny"}
dat=read_csv("phco2.csv")
tbl_df(dat)
```

# Plot patterns among subjects
```{r, echo=F}
ggplot(dat,aes(x=co2,y=ph,group=id))+geom_line()

```

## Set of models to be fit
1. $$Y_{ijk} = \beta_0  + \beta_1 X_{ijk}+\beta_2 X^2_{ijk}+e_{ijk}$$
2. $$Y_{ijk} = \beta_0  + \beta_{0i}+ \beta_1 X_{ijk}+\beta_2 X^2_{ijk}+e_{ijk}$$
3. $$Y_{ijk} = \beta_0  + \beta_{0i} + \beta_{0ij}+ \beta_1 X_{ijk}+\beta_2 X^2_{ijk}+e_{ijk}$$
4. $$Y_{ijk} = \beta_0  + \beta_{0i}+ (\beta_1+\beta_{1i}) X_{ijk}+\beta_2 X^2_{ijk}+e_{ijk}$$

## Model 1
```{r mod1, size="tiny"}
# Make a co2^2 variable
dat <- dat %>% mutate(co22=co2^2)
#  OLS
lm_1 <- glm(ph ~ co2+co22, data=dat,family="gaussian")
summary(lm_1)
# Residual SD
sigma(lm_1)
# AIC
AIC(lm_1)
```

## Estimates of changes in mean for different changes in CO2
```{r post1}
# Note, if co2 is x, then co22 is x^2
comps=rbind(c(0,3,9),c(0,5,25),c(0,7,49))
labels=c("3 vs 0", "5 vs. 0", "7 vs. 0")
source("glm_post_estimate.R")
post1 <- glm.post.estimate(lm_1,comps,labs=labels)
post1
```


## Model 2
```{r mod2, size="tiny"}
lme_2 <- lmer(ph ~ co2+co22 + (1|id), data=dat)
summary(lme_2)
AIC(lme_2)
```

## Model 2 Stats
```{r post2a, size="tiny"} 
# ICC - var(b0i)/(var(b0i)+var(eij))
icc(lme_2)
AIC(lme_2)
# Now do by grabbing relevant objects
rand_vars2 <- re_var(lme_2)
rand_vars2
rand_vars2[2]/(rand_vars2[1]+rand_vars2[2])
```


## Estimates of changes in mean for different changes in CO2
```{r post2b}
source("glmer_post_estimate.R")
post2 <- glmer.post.estimate(lme_2,comps,labs=labels,
                            exponentiate = F)
post2
```

## Model 3
```{r mod3, size="tiny"}
lme_3 <- lmer(ph ~ co2+co22 + (1|id/visit), data=dat)
summary(lme_3)
AIC(lme_3)
```

## Model 3 Stats
```{r post3a, size="tiny"} 
# ICC
icc(lme_3)
# Now do by grabbing relevant objects
rand_vars3 <- re_var(lme_3)
rand_vars3
# ICC id
rand_vars3[3]/(sum(rand_vars3))
# ICC visit in id
rand_vars3[2]/(sum(rand_vars3))
```

## Estimates of changes in mean for different changes in CO2
```{r post3b}
post3 <- glmer.post.estimate(lme_3,comps,
                             labs=labels,exponentiate = F)
post3
```

## Model 4
```{r mod4, size="tiny"}
lme_4 <- lmer(ph ~ co2+(co2|id)+co22, data=dat)
summary(lme_4)
AIC(lme_4)
```

## Model 4 Stats
```{r post4a, size="tiny"} 
# ICC
icc(lme_4)
# Now do by grabbing relevant objects
rand_vars4 <- re_var(lme_4)
rand_vars4
# ICC id at Xijk=0
rand_vars4[2]/(rand_vars4[1]+rand_vars4[2])
```

## Estimates of changes in mean for different changes in CO2
```{r post4b}
# Note, if co2 is x, then co22 is x^2
post4 <- glmer.post.estimate(lme_4,comps,
                             labs=labels,exponentiate = F)
post4
```

## Summary stats

\begin{center}
\tiny{
\begin{table}
\caption{\textsc{Estimates(SE) and Akaike's Information Criterion (AIC) of four hierarchical linear mixed effects models for pH versus $C0_2$}}
\label{ch7tab.phvsCO2}
\vspace{0.2in}
\begin{tabular}{ccccccccc}
& \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2} & \multicolumn{2}{c}{Model 3} & \multicolumn{2}{c}{Model 4} \\
\hline
 Parameter & Est & SE & Est & SE & Est & SE & Est & SE   \\
\hline
$\beta_0$ &  7.6 &   .021 &    7.6 &   .021 &    7.6 &    .021 &    7.6 &   .026  \\
$\beta_1$ & -.14 &   .014 &   -.14 &   .008 &   -.14 &  .0080 &   -.14 &  .0076 \\
$\beta_2$ & .0073 &  .0019 &  .0074 &  .0011 &  .0073 &  .0011 &  .0073 &   .0010 \\
\hline
$\sigma_{\beta_{0i}}$  &     &  &   .071 &   &   .071 &    &   .097 &    \\ \
$\sigma_{\beta_{0ij}}$ &   &       &       &       &   .021 &    &       &       \\
$\sigma_{\beta_{1i}}$ &   &       &       &       &       &      &  .0065 &  \\
$\sigma_e$ &  .088 &   &   .051 &   &   .048 &   &   .047 &   \\
\hline
$AIC$ & -136 &       &   -143 &       &   -142 &       &   -147 &       \\ 
\end{tabular} \\[0.5ex]
\end{table}
}
\end{center}

## Model vs. Association of Interest
```{r sumfig, echo=F}
# Clunky so don't take as inspiration for your code
mod <- c(1,as.numeric(substr(post1[2,2],1,6)),as.numeric(substr(post1[2,1],1,6)),as.numeric(substr(post1[2,2],9,14)))
mod <- rbind(mod,c(2,as.numeric(substr(post2[2,2],1,6)),as.numeric(substr(post2[2,1],1,6)),as.numeric(substr(post2[2,2],9,14))))
mod <- rbind(mod,c(3,as.numeric(substr(post3[2,2],1,6)),as.numeric(substr(post3[2,1],1,6)),as.numeric(substr(post3[2,2],9,14))))
mod <- rbind(mod,c(4,as.numeric(substr(post4[2,2],1,6)),as.numeric(substr(post4[2,1],1,6)),as.numeric(substr(post4[2,2],9,14))))
mod <- data.frame(mod)
names(mod) <- c("model","Y0","Y1","Y2")
ggplot(mod,aes(x=model,y=Y0))+geom_line(linetype = "dashed")+geom_line(aes(x=model, y=Y1))+geom_line(aes(x=model, y=Y2),linetype = "dashed")+ ylab("E(Yijk | Xijk=5)-E(Yijk | Xijk=0)")+ggtitle('Estimate and 95 % CI (dashed lines)')

```

