---
title: 'Longitudinal Data: Repeated Measures'
subtitle: 'More GEE'
author: Alan Hubbard
date: Oct 23, 2018
output: 
  beamer_presentation:
    slide_level: 2   
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
# This bit of code lets you set the font size
# in heading of code junk when writing to beamer.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
# Libraries
library(dplyr)
library(tidyverse)
library(tidyr)
library(gee)
library(cowplot)
library(data.table)
```


## Intro
- We will examine 3 different analyses all from the CD4/HIV data.
- For each, we will compare
    + standard linear regression ignoring repeated measures
    + standard linear regression using GEE to adjust inference robustly to account for repeated measures (GEE with independence working correlation, robust variance)
    + GEE with exchangeable working correlation, naive variance
    + GEE with exchangealbe working correlation, robust variance
- Purpose is to compare
    + coefficient estimates comparing independence and different correlation model, and
    + inference comparing robust versus naive (assuming model is correct) versus robust (agnostic correlation model)

# Binary time-independent variable (balanced data)

## Read in data
```{r prelim1, size="tiny"}
dat1=read_csv("deeks_ex1.csv")
tbl_df(dat1)
```

## Independence Working Correlation
```{r ex1, size="tiny"}
# independence
gee_ind=gee_fit <- gee(cd4 ~ binage, id, data=dat1,family=gaussian, corstr = "independence")
# Make easier to read summary
ss1 <- data.frame(summary(gee_fit)$coefficients)
ss1 = data.frame(ss1,pvalue=2*(1-pnorm(abs(ss1[,5]))))
round(ss1,4)
```

## Exchangeable
```{r ex1b, size="tiny"}
gee_exc <- gee(cd4 ~ binage, id, data=dat1,family=gaussian, corstr = "exchangeable")
# Make easier to read summary
ss2 <- data.frame(summary(gee_exc)$coefficients)
ss2 = data.frame(ss2,pvalue=2*(1-pnorm(abs(ss2[,5]))))
round(ss2,4)
```    

# Binary time-dependent variable (balanced data)
## Read in data
```{r prelim2, size="tiny"}
dat2=read_csv("deeks_ex2.csv")
tbl_df(dat2)
```

## Independence Working Correlation
```{r ex2, size="tiny"}
# independence
gee_ind <- gee(cd4 ~ medvl, id, data=dat2,family=gaussian, corstr = "independence")
# Make easier to read summary
ss1 <- data.frame(summary(gee_ind)$coefficients)
ss1 = data.frame(ss1,pvalue=2*(1-pnorm(abs(ss1[,5]))))
round(ss1,4)
```

## Exchangeable
```{r ex2b, size="tiny"}
gee_exc <- gee(cd4 ~ medvl, id, data=dat2,family=gaussian, corstr = "exchangeable")
# Make easier to read summary
ss2 <- data.frame(summary(gee_exc)$coefficients)
ss2 = data.frame(ss2,pvalue=2*(1-pnorm(abs(ss2[,5]))))
round(ss2,4)

```    


# Continuous time-dependent variables (very unbalanced data)

## Read in data
```{r prelim3, size="tiny"}
dat3=read_csv("deeks_ex3.csv")
tbl_df(dat3)
```

## Independence Working Correlation

```{r ex3, size="tiny"}
# independence
gee_ind <- gee(cd4 ~ logvlbase+logvlchange, id, data=dat3,na.action=na.omit, family=gaussian, corstr = "independence")
# Make easier to read summary
ss1 <- data.frame(summary(gee_ind)$coefficients)
ss1 = data.frame(ss1,pvalue=2*(1-pnorm(abs(ss1[,5]))))
round(ss1,4)
```

## Exchangeable
```{r ex3b, size="tiny"}
gee_exc <- gee(cd4 ~ logvlbase+logvlchange, id, data=dat3,na.action=na.omit, family=gaussian, corstr = "exchangeable")
# Make easier to read summary
ss2 <- data.frame(summary(gee_exc)$coefficients)
ss2 = data.frame(ss2,pvalue=2*(1-pnorm(abs(ss2[,5]))))
round(ss2,4)
```    


## Why large differences in estimates returned by same linear model, but different working correlation models
- Exchangeable will tends to give more influence to subjects with fewer observations relative to independence working correlation.
- Below, we use code from chapter 2 using tge data.table package to estimate the slope of CD4 versus logVL by id.
- First, make a variable that is the log(VL), which is simply baseline value plus change.
- We look at the subset of id's that have at least 10 repeated observations.
- We then plot the distribution of slopes.



## Histogram of estimated slopes
```{r slopesbyid, size="tiny", echo=F}
# Data.table again
dat3_expl <- dat3 %>% group_by(id) %>% mutate(logVLt = logvlbase+logvlchange) %>% mutate(num_dist = n_distinct(logVLt))
dat3_sht <- data.table(subset(dat3_expl, num_dist>10))
out <- as.tibble(dat3_sht[,as.list(coef(lm(cd4 ~ logVLt,na.action = na.omit))),by=.(id)])
p1 <- ggplot(out,aes(x=logVLt))+  geom_histogram(bins=15,col="black",fill="white")+
  ggtitle("Distribution of estimated slopes of logVL vs. CD4")+
  labs(x="slope")+ 
  theme(plot.title = element_text(size=12)) +
  geom_vline(xintercept=mean(out$logVLt,na.rm = T))
p1
```

## Results
- See that the mean of the slopes is close to that found in the exchangeable gee analysis above.
- Implies that the exchangeable working correlation gives something closer to equal weights to individuals than the independence.
-Independence gives more weight to individuals with more observations.





